---
title: "functions.Rmd"
author: "Matthijs Gielen"
date: "2025-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This document provides an overview of the functions defined in the `functions.R` script. This script is intended to be used in conjuction with the main script **`AlienDetective.R`**.  

The `functions.R` script includes the following functions:  
- `fetch_gbif_data()`  
- `is_on_land()`  
- `move_to_sea()`  
- `calculate.distances()`  

---  

# Function `fetch_gbif_data()`  

This `fetch_gbif_data()` function retrieves occurrence data from GBIF for a specified species.  
## Arguments  

The function allows customization through several arguments:  
- `species`: The scientific name of the species.  
- `hasCoordinate`: If `TRUE`, only return records with coordinates.  
- `continent`: Filters results by continent.  
- `basisOfRecord`: Specifies desired record types.  
- `fields`: Column names to include in the result.  
- `limit`: Maximum number of results to fetch.  
- `output_dir`: Directory where output (if saved) would be stored.  

>**Note:** Refer to the  [`occ_search`](https://docs.ropensci.org/rgbif/reference/occ_data.html?q=occ_search) documentation for full details on accepted parameter values.  

## Function Definition

```{r}
fetch_gbif_data <- function(species,
                            hasCoordinate = TRUE,
                            continent = "europe",
                            basisOfRecord = c("OBSERVATION", "MACHINE_OBSERVATION", "HUMAN_OBSERVATION", "MATERIAL_SAMPLE", "LIVING_SPECIMEN", "OCCURRENCE"),
                            fields = c("decimalLatitude", "decimalLongitude", "year", "month", "country"),
                            limit = 10000,
                            output_dir) {
  # Actual code in function (see next blocks)
}
```

>The logic of the function is split into sections to explain each step of the function.  

## Step 1: Download GBIF Data  

This step uses the parameters above to request occurrence data from the GBIF API via the `rgbif::occ_search()` function.  

```{r}
data_list <- rgbif::occ_search(scientificName = species,
                               hasCoordinate = TRUE,
                               continent = continent,
                               basisOfRecord = basisOfRecord,
                               fields = fields,
                               limit = limit)
```

## Step 2: Prepare and Combine Records  

We initialize an empty data frame, then iterate over the downloaded records to:  
- Skip empty or missing entries  
- Fill in any missing columns with `NA`  
- Combine all valid entries into one final data frame (`res`)  

```{r}
# Initialise the empty data frame
res <- data.frame()

# Loop over each data frame in the list
for (i in seq_along(data_list)) {
  # Skip the basis of record(s) that have no data
  if (is.null(data_list[[i]]$data)) {
    next
  } else if (nrow(data_list[[i]]$data) == 0) {
    next
  }
  
  # Replace missing values with "NA"
  missing_columns <- setdiff(fields, colnames(data_list[[i]]$data))
  if (length(missing_columns) > 0) {
    for (col in missing_columns) {
      data_list[[i]]$data[col] <- NA
    }
  }
  
  # Concatenate into 1 data frame
  data_list[[i]]$data$basisOfRecord <- names(data_list)[i]
  res <- rbind(res, data_list[[i]]$data)
}
```

## Step 3: Final Formatting and Return  

This step formats the final data frame and filters out incomplete records.  

```{r}
if (nrow(res) > 0) {
  # Reorder columns by order in fields argument
  res <- res[, c(fields, "basisOfRecord")]
  
  # Rename latitude & longitude columns
  colnames(res)[colnames(res) == "decimalLatitude"] <- "latitude"
  colnames(res)[colnames(res) == "decimalLongitude"] <- "longitude"
  
  # Remove occurrences where longitude or latitude is "NA"
  res <- res[!is.na(res$latitude) & !is.na(res$longitude),]
  return(res)
} else {
  error_message <- paste0("No GBIF records found for species \"", species, "\"")
  message(error_message)
  return(NULL)
}
```

